\sysfull

### 2017-11-15 paper/paging.tex
paging

\sys paging scheme partitions the non-volatile memory address space into pages of {\em protected data}. As they are accessed by a task, pages of protected data are swapped into a {\em page buffer} that is stored in volatile memory and holds a single page. When a modified page is swapped out, it is buffered in a dedicated commit buffer for later. \sys uses two-phase commit to write buffered, modified pages back to their original memory locations atomically, ensuring data consistency even with power failures. \sys's page-based memory model allows moving large blocks of data between non-volatile and volatile memory, which \sys does efficiently using common,
hardware DMA block-copy operations, unlike prior work~\cite{chain,alpaca} that moves one variable at a time. Moreover, \sys's paging and buffering scheme is a key enabler of task coalescing, which we describe in Section~\ref{sec:task_coalescing}. Paging simplifies coalescing because deferring a commit of paged data at a task boundary entails leaving paged data buffered only, not requiring additional state management logic at task boundaries.

@@

\sys virtualizes the protected, non-volatile data that a task can access through a modified memory interface. The interface includes two operations: \texttt{RVAR(var)}, which reads the value of the memory location {\tt var} and \texttt{WVAR(var,val)}, which writes the value {\tt val} into the memory location {\tt var}. {\tt RVAR} and {\tt WVAR} operations translate a variable's physical address in non-volatile memory and translates it into a \emph{virtual address} in the page buffer \texttt{pageBuf} in volatile memory.  In \sys, a virtual address is composed of a \emph{page tag} and a \emph{page offset}. The page tag identifies in which page of non-volatile memory the memory location resides and the page offset identifies a particular byte of that page.

@@

\begin{algorithm}[t]
	\caption{\texttt{RVAR(var)} pseudo-code}
	\label{algo:rwar}
	\scriptsize
	%\small
	\begin{algorithmic}[1]
		\State $\texttt{tag}\leftarrow \texttt{getTag(var)}$ 
		\If { \texttt{tag} != \texttt{CrntPagTag} }	\Comment{Check if the page is in the page buffer}
		\State	\texttt{PageFault(tag)} \Comment{Page is not in the page buffer, bring it}
		\EndIf
				\State $\texttt{offset}\leftarrow \texttt{getOffset(var)}$ 		
		\State \texttt{return pageBuf[offset]}  \Comment{Return directly from page buffer}
	\end{algorithmic}
\end{algorithm}

After address translation, \sys attempts to access the protected variable's location in the volatile paging buffer. \sys keeps track of the page tag for the page currently resident in the paging buffer in {\tt CrntPagTag}, which is stored in non-volatile memory. When a task accesses a variable, it compares the variable's page tag to {\tt CrntPagTag}.  If an accessed variable's page is equal to {\tt CrntPagTag} then the accessed variable's page is resident in the page buffer. The executing memory operation indexes the page buffer using the variable's page offset and manipulates that byte (i.e., either reading or writing). If the accessed variable's page tag is not equal to {\tt CrntPagTag} then the variable's page is not resident in the page buffer; the operation incurs a {\em page fault}. At a page fault, \sys must swap out the page that is resident in the page buffer and swap in the the variable's page.

Algorithm~\ref{algo:rwar} shows pseudo-code of \texttt{RVAR} and we omit a listing for {\tt WVAR} as it is very similar except it writes the memory location and returns nothing. If the page tag of the variable \texttt{var} and \texttt{CrntPagTag} are not equal (e.g. Lines 2--3), there is a page fault that requires a new page to be swapped in to the page buffer.

@@

\begin{algorithm}[t]
	\caption{\texttt{PageFault(tag)} pseudo-code}
	\label{algo:pagefault}
	\scriptsize
	%\small
	\begin{algorithmic}[1]
		\If {\texttt{isDirty(CrntPagTag)} }	\Comment{Check if the active page is modified}
		\State \texttt{pageCommit(pagesTemp)} \Comment{Phase I commit the active page}
		\EndIf
		\If {\texttt{isDirty(tag)}} \Comment{Check if the requested page has been modified}
		\State {\texttt{getPage(tag,pagesTemp)}} \Comment{Get page from the temporary buffer}
		\Else
		\State \texttt{getPage(tag,pagesOrg)} \Comment{Get page from the original location}
		\EndIf 
		\State \texttt{CrntPagTag} = \texttt{tag} \Comment{Update tag}
	\end{algorithmic}
\end{algorithm}

A page fault on a memory access requires \sys to swap out whatever page of data is in the page buffer, preserving updates made to that page, and to copy the accessed page of protected data into the page buffer. Algorithm~\ref{algo:pagefault} shows the how \sys handles a page fault. If both the page being swapped out and the page being swapped in are not {\em dirty}---which means that they do not contain bytes modified by the executing task---then handling a page fault simply requires copying the accessed page into the page buffer. If either the {\em resident} page being swapped out or the {\em accessed} page being swapped in is dirty, \sys must take extra steps when handling a page fault.

\paragraph{Dirty resident page.} If a task modified any byte in the resident page (i.e.,  using \texttt{WVAR}), then the page is dirty. Before swapping in the accessed page, \sys must first preserve the updated values in the dirty page so that they can eventually be committed back to their locations in memory when the task ends. To preserve dirty values, \sys copies the resident page to its dedicated memory region called \texttt{pagesTemp}, which is acts as a double buffer for each page of protected memory. Dirty pages copied into {\tt pagesTemp} during a task remain there until the task completes, at which point \sys commits the modified pages from {\tt pagesTemp} back to memory.

\paragraph{Dirty accessed page.} If an accessed page is modified, \sys should swap it into the page buffer from its location in {\tt pagesTemp}, rather than swapping it in from its original location in memory. As Algorithm~\ref{algo:pagefault} shows in Line 3, \sys checks if the requested page was previously modified by the executing task. To track whether a page is modified, \sys stores a dirty bit with the page that is set when a task modifies the page. If an accessed page's dirty bit is set, the page must be swapped in from \texttt{pagesTemp} rather than its original memory location (Line 4), furnishing the task with the page's modified values. If the accessed page is not dirty, \sys swaps it in from its original location in memory (Line 6). After swapping the page in, \sys sets \texttt{CrntPagTag} to the tag of the accessed page (Line 7). \sys uses DMA block copies for all page swapping operations, leveraging the efficiency of hardware support.

@@

Figure~\ref{fig:volatile-buffer} illustrates the execution of a simple task using \sys's paging mechanism. The task manipulates the protected variables {\tt x}, {\tt y}, and {\tt z}. The variables {\tt y} and {\tt z} are in the same page and they are currently in the volatile page buffer---on accessing {\tt y} and {\tt z}, the task needs only to access \texttt{pageBuf}. Since variable \texttt{x} is kept in a different page in \texttt{pagesOrg}, accessing it creates a page fault. At this point, \sys swaps the active page in \texttt{pageBuf} by committing it into the non-volatile {\tt pagesTemp} region and copying the corresponding page from \texttt{pagesOrg} to \texttt{pageBuf}. When the task completes, \sys sets the commit bit, copies the dirty page from \texttt{pagesTmp} to its original location in \texttt{pagesOrg} and unsets the commit bit.

>>>>

\sys paging scheme partitions the non-volatile memory address space into pages of {\em protected data}. As they are accessed by a task, pages of protected data are swapped into a {\em page buffer} \texttt{pagesBuf} that is stored in volatile memory and holds a pre-determined number of pages. When any modified page in the page buffer is swapped out, it is buffered in a dedicated {\em commit buffer} \texttt{pagesTemp} for later. \sys uses two-phase commit to write buffered, modified pages back to their original memory locations atomically, ensuring data consistency even with power failures. \sys's page-based memory model allows moving large blocks of data between non-volatile and volatile memory, which \sys does efficiently using common,
hardware DMA block-copy operations, unlike prior work~\cite{chain,alpaca} that moves one variable at a time. Moreover, \sys's paging and buffering scheme is a key enabler of task coalescing, which we describe in Section~\ref{sec:task_coalescing}. Paging simplifies coalescing because deferring a commit of paged data at a task boundary entails leaving paged data buffered only, not requiring additional state management logic at task boundaries.

@@

\sys virtualizes the protected, non-volatile data that a task can access through a modified memory interface. The interface includes two operations: \texttt{RVAR(var)}, which reads the value of the memory location {\tt var} and \texttt{WVAR(var,val)}, which writes the value {\tt val} into the memory location {\tt var}. {\tt RVAR} and {\tt WVAR} operations translate a variable's physical address in non-volatile memory and translates it into a \emph{virtual address} in the page buffer in volatile memory.  In \sys, a virtual address is composed of a \emph{page tag} and a \emph{page offset}. The page tag identifies in which page of non-volatile memory the memory location resides and the page offset identifies a particular byte of that page.

@@

\begin{algorithm}[t]
	\caption{\texttt{RVAR(var)} pseudo-code}
	\label{algo:rwar}
	\scriptsize
	%\small
	\begin{algorithmic}[1]
		\State $\texttt{tag}\leftarrow \texttt{getTag(var)}$ 
		\If { \texttt{tag} is not found in \texttt{pagesBuf} }	\Comment{Check if the page is in the page buffer}
		\State	\texttt{PageFault(tag)} \Comment{Page is not in the page buffer, bring it}
		\Else 
		\State \texttt{idx} $\leftarrow$ \texttt{getPageIndex()} \Comment{Get the page index in the page buffer}
		\EndIf
				\State $\texttt{offset}\leftarrow \texttt{getOffset(var)}$ 		
		\State \texttt{return pagesBuf[idx][offset]}  \Comment{Return directly from page buffer}
	\end{algorithmic}
\end{algorithm}

After address translation, \sys attempts to access the protected variable's location in the volatile paging buffer. \sys keeps track of the page tags for the pages currently resident in the paging buffer. When a task accesses a variable, it compares the variable's page tag to tags of the pages in {\tt pagesBuf}.  If the accessed variable's page is resident in the page buffer, then the tag of one of the pages in this buffer is equal to the accessed variable's page tag. The executing memory operation indexes the page buffer using the variable's page offset and manipulates that byte (i.e., either reading or writing). If the accessed variable's page tag is not found then the variable's page is not resident in the page buffer; the operation incurs a {\em page fault}. At a page fault, \sys must swap out one of the pages that is resident in the page buffer and swap in the variable's page.

Algorithm~\ref{algo:rwar} shows pseudo-code of \texttt{RVAR} and we omit a listing for {\tt WVAR} as it is very similar except it writes the memory location and returns nothing. If the page tag of the variable \texttt{var} is not found in \texttt{pagesBuf} (e.g. Lines 2--3), there is a page fault that requires a new page to be swapped in to the page buffer.

@@

\begin{algorithm}[t]
	\caption{\texttt{PageFault(tag)} pseudo-code}
	\label{algo:pagefault}
	\scriptsize
	%\small
	\begin{algorithmic}[1]
		\State \texttt{VictimTag} $\leftarrow$ \texttt{selectVictimPage()} \Comment{Select a page to swap out}
		\If {\texttt{isDirty(VictimTag)} }	\Comment{Check if the victim page is modified}
		\State \texttt{pageCommit(VictimTag,pagesTemp)} \Comment{Phase I commit the victim page}
		\EndIf
		\If {\texttt{isDirty(tag)}} \Comment{Check if the requested page has been modified}
		\State {\texttt{getPage(tag,pagesTemp)}} \Comment{Get page from the temporary buffer}
		\Else
		\State \texttt{getPage(tag,pagesOrg)} \Comment{Get page from the original location}
		\EndIf 
	\end{algorithmic}
\end{algorithm}

A page fault on a memory access requires \sys to swap out one of the pages in the page buffer, i.e. \emph{victim page}, preserving updates made to that page, and to copy the accessed page of protected data into the page buffer. Algorithm~\ref{algo:pagefault} shows the how \sys handles a page fault. In our current implementation, we used a \emph{first-in-first-out} page replacement algorithm to select the victim page in \texttt{pagesBuf} (Line 1). If both the page being swapped out and the page being swapped in are not {\em dirty}---which means that they do not contain bytes modified by the executing task---then handling a page fault simply requires copying the accessed page into the page buffer. If either the {\em victim} page being swapped out or the {\em accessed} page being swapped in is dirty, \sys must take extra steps when handling a page fault.

\paragraph{Dirty victim page.} If a task modified any byte in the victim page (i.e.,  using \texttt{WVAR}), then the page is dirty. Before swapping in the accessed page, \sys must first preserve the updated values in the dirty page so that they can eventually be committed back to their locations in memory when the task ends. To preserve dirty values, \sys copies the victim page to its dedicated memory region called \texttt{pagesTemp}, which is acts as a double buffer for each page of protected memory. Dirty pages copied into {\tt pagesTemp} during a task remain there until the task completes, at which point \sys commits the modified pages from {\tt pagesTemp} back to memory (Lines 2--3).

\paragraph{Dirty accessed page.} If an accessed page is modified, \sys should swap it into the page buffer from its location in {\tt pagesTemp}, rather than swapping it in from its original location in memory. As Algorithm~\ref{algo:pagefault} shows in Line 4, \sys checks if the requested page was previously modified by the executing task. To track whether a page is modified, \sys stores a dirty bit with the page that is set when a task modifies the page. If an accessed page's dirty bit is set, the page must be swapped in from \texttt{pagesTemp} rather than its original memory location (Line 5), furnishing the task with the page's modified values. If the accessed page is not dirty, \sys swaps it in from its original location in memory (Line 7). \sys uses DMA block copies for all page swapping operations, leveraging the efficiency of hardware support.

@@

Figure~\ref{fig:volatile-buffer} illustrates the execution of a simple task using \sys's paging mechanism. The task manipulates the protected variables {\tt x}, {\tt y}, and {\tt z}. The variables {\tt y} and {\tt z} are in the same page and they are currently in the volatile page buffer---on accessing {\tt y} and {\tt z}, the task needs only to access \texttt{pageBuf}. Since variable \texttt{x} is kept in a different page in \texttt{pagesOrg}, accessing it creates a page fault. At this point, \sys swaps the active page in \texttt{pagesBuf} by committing it into the non-volatile {\tt pagesTemp} region and copying the corresponding page from \texttt{pagesOrg} to \texttt{pageBuf}. When the task completes, \sys sets the commit bit, copies the dirty page from \texttt{pagesTmp} to its original location in \texttt{pagesOrg} and unsets the commit bit.
